Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-03-05 sources/bmc_template_submission/bmc_article.tex
more results

An online literature review in June 2020 (Google scholar search for "ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data publicly available \cite{gottlieb2005molecular, yao2020phylogeny}.
\cite{gottlieb2005molecular} made original tree and alignment available in TreeBASE
(study 1091 \cite{treebase1091}). This tree (Gottlieb2005 tree from now on) samples 41 species, is available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The most recent \textit{Ilex} tree \cite{yao2020phylogeny} (Yao2020 tree from now on) is available in
OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the Yao2020 tree is the best sampled phylogeny yet available for
the genus \textiy{Ilex}.

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment from \cite{gottlieb2005molecular}, using a local
GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
%% <!-- MTH: you probably need some details about the hardware, given the fact that you are discussing running times -->
The updated Gottlieb2005 tree (Figure 2) displays
all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional \textit{Ilex} taxa. The best RaxML tree is 99\% resolved, with 25\% of nodes
with bootstrap support $<$ 0.1 and 48\% nodes with bootstrap support $>$ 0.75.
A large portion of internal branches are negligibly small, with 30 branches
$<$ 0.00001 substitution rate units, from which only 9 have a bootstrap support $>$ 0.75
(Figure 2).
The Yao2020 tree \cite{yao2020phylogeny} also contains all 41 distinct taxa
sampled in the Gottlieb2005 tree \cite{gottlieb2005molecular},
while contributing phylogenetic data to 134 additional \textit{Ilex} taxa. From these,
67 taxa are also in the Physcraper updated Gottlieb2005 tree \cite{gottlieb2005molecular}.
While the Yao2020 tree \cite{yao2020phylogeny} was also constructed using
ITS as a marker, their GenBank data is not released yet. Hence, Physcraper was unable
to incorporate 68 taxa that are only on the Yao2020 tree because the DNA data is unavailable.
We also note that Physcraper incorporates 18 \textit{Ilex} taxa that are not in the Yao2020
tree \cite{yao2020phylogeny}. These taxa appear nested among other \textit{Ilex} species
and visual inspection of the DNA sequences suggests they are correctly assigned as \textit{Ilex}.
The ITS alignment that underlies the Yao2020 tree was constructed without any tool to
scrape GenBank \cite{yao2020phylogeny}, which could explain why Physcraper was able to
incorporate these 18 additional \textit{Ilex} taxa.

>>>>

An online literature review in June 2020 (Google scholar search for "ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data publicly available \cite{gottlieb2005molecular, yao2020phylogeny}.
\cite{gottlieb2005molecular} made original tree and alignment available in TreeBASE
(study 1091 \cite{treebase1091}). This tree (Gottlieb2005 tree from now on) samples 41 species, is available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The most recent \textit{Ilex} tree \cite{yao2020phylogeny} (Yao2020 tree from now on) is available in
OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the Yao2020 tree \cite{yao2020phylogeny} is the best sampled phylogeny available for
the genus \textit{Ilex}.

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment that was used to construct the tree from \cite{gottlieb2005molecular},
using a local GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
%% <!-- MTH: you probably need some details about the hardware, given the fact that you are discussing running times -->
The Gottlieb2005 tree \cite{gottlieb2005molecular} updated using Physcraper (Figure 2;
Physcraper updated tree from now on) displays all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional \textit{Ilex} taxa. The best ML tree from the RAxML analysis is 99\%
resolved, with 25\% of nodes with bootstrap support $<$ 0.1 and 48\% nodes with
bootstrap support $>$ 0.75.
A large portion of internal branches are negligibly small, with 30 branches
$<$ 0.00001 substitution rate units, from which only 9 have a bootstrap support $>$ 0.75
(Figure 2).
As comparison with the Physraper updated tree, the Yao2020 tree \cite{yao2020phylogeny}
also contains all 41 distinct taxa
sampled in the Gottlieb2005 tree \cite{gottlieb2005molecular},
while contributing phylogenetic data to 134 additional \textit{Ilex} taxa. From these,
67 taxa are also in the Physcraper updated tree.
While the Yao2020 tree \cite{yao2020phylogeny} was also constructed using
ITS as a marker, their GenBank data is not released yet. Hence, Physcraper was unable
to incorporate 68 taxa that are only on the Yao2020 tree because the DNA data is unavailable.
We also note that Physcraper incorporates 18 \textit{Ilex} taxa that are not in the Yao2020
tree \cite{yao2020phylogeny}. These taxa appear nested among other \textit{Ilex} species
and visual inspection of the DNA sequences suggests they are correctly assigned as \textit{Ilex}.
The ITS alignment that underlies the Yao2020 tree was constructed without any tool to
scrape GenBank \cite{yao2020phylogeny}, which could explain why Physcraper was able to
incorporate these 18 additional \textit{Ilex} taxa in the updated tree (Figure 2).
